---
title: "Final Project Data 606 Statistical analysis"
author: "Mehreen Ali Gillani"
output: html_document
date: "2025-11-26"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Statistical Analysis of U.S. Domestic Air Travel (Jan–July 2025)

##  Project Overview

This statistical analysis examines seasonal and operational patterns in U.S. domestic air travel using the T‑100 Domestic Market (U.S. Carriers) dataset from January to July 2025. The study employs hypothesis testing to investigate two key dimensions of air travel dynamics.

##  Research Questions & Methodology

### 1. Seasonal Flight Distribution

We tested whether flight frequencies are uniformly distributed across months using a chi‑square goodness‑of‑fit test. Our analysis provides strong evidence against equal monthly distribution (χ² = 155.54, df = 6, p ≈ 0), indicating airlines strategically adjust schedules in response to seasonal travel patterns.

### 2. Carrier‑Type Performance

We evaluated differences in flight frequencies and passenger volumes across three carrier classifications—major, national, and regional—using:

Chi‑square tests for flight frequency comparisons
ANOVA for passenger volume analysis
Both tests yielded statistically significant results (p ≈ 0), rejecting the null hypothesis of equal performance across carrier types. Major carriers dominate operations, accounting for 61% of flights and 57.6% of passengers.

### Analytical Framework

This report details the inferential procedures, test assumptions, and practical interpretations of these findings, providing evidence‑based insights into seasonal scheduling strategies and carrier‑type operational differentiation in the U.S. domestic aviation market.


```{r}
# import Libraries

library(GGally)
library(scales)
library(tidyverse)


```

**Read Flights**

```{r read_file}
url <- "https://raw.githubusercontent.com/mehreengillani/Final_project_Data606/refs/heads/main/flights_data_clean.csv"
flights <- read_csv(url)

# Verify the loaded data
head(flights)

```

### Research Questions

**1. Seasonal Flight Operations:** How do flight frequencies vary across
months? Statistical Framework:

### Null Hypothesis (H₀): There is no significant difference in mean flight
frequencies across months (μ_jan = μ_feb = μ_mar = μ_apr = μ_may = μ_jun
= μ_jul)

###Alternative Hypothesis (H₁): Mean flight frequencies differ
significantly across months, with summer months (June-July) showing
elevated frequencies driven by seasonal travel patterns \* 

**Response:**
flight_count (Numerical) 

**Explanatory:** month (Categorical)

```{r}
# Create summary data for monthly analysis
monthly_summary <- flights %>%
  group_by(month) %>%
  summarise(flight_count = n()) %>%
  mutate(month_name = factor(c("January", "February", "March", "April", "May", "June", "July")[month],
                            levels = c("January", "February", "March", "April", "May", "June", "July")))

# View the summary data
print(monthly_summary)
```

```{r}
# Basic statistics
cat("\n=== SUMMARY STATISTICS ===\n")
cat("Total flights:", format(sum(monthly_summary$flight_count), big.mark = ","), "\n")
cat("Average monthly flights:", format(mean(monthly_summary$flight_count), big.mark = ","), "\n")
cat("Month with highest flights:", monthly_summary$month_name[which.max(monthly_summary$flight_count)], "\n")
cat("Month with lowest flights:", monthly_summary$month_name[which.min(monthly_summary$flight_count)], "\n")

# Calculate percentage change from lowest to highest month
lowest_count <- min(monthly_summary$flight_count)
highest_count <- max(monthly_summary$flight_count)
percentage_increase <- ((highest_count - lowest_count) / lowest_count) * 100
cat("Seasonal increase from lowest to highest month:", round(percentage_increase, 1), "%\n")
```

```{r}
# Create the main visualization - Bar chart
ggplot(monthly_summary, aes(x = month_name, y = flight_count, fill = flight_count)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = format(flight_count, big.mark = ",")), vjust = -0.5, size = 3.5) +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Flight Count") +
  labs(title = "Flight Frequencies by Month",
       subtitle = "Analysis of Seasonal Patterns in Flight Operations",
       x = "Month",
       y = "Number of Flights",
       caption = paste("Total flights analyzed:", format(sum(monthly_summary$flight_count), big.mark = ","))) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.1))) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(face = "bold"))


```

```{r}
# Calculate month-over-month changes
monthly_changes <- monthly_summary %>%
  mutate(previous_month = lag(flight_count),
         mom_change = flight_count - previous_month,
         mom_percentage = ((flight_count - previous_month) / previous_month) * 100)
cat("\n=== MONTH-OVER-MONTH CHANGES ===\n")
print(monthly_changes %>% select(month_name, flight_count, mom_change, mom_percentage))
```


```{r}
# Chi-Square Test: Are flights distributed equally across months?
cat("\n=== CHI-SQUARE GOODNESS OF FIT TEST ===\n")

monthly_counts <- flights %>%
  count(month_name) %>%
  arrange(match(month_name, c("jan", "feb", "mar", "apr", "may", "jun", "jul")))
#print(monthly_counts$n)
# Expected frequencies (if equal distribution)
expected <- rep(sum(monthly_counts$n) / 7, 7)
print(expected)
chi_test <- chisq.test(monthly_counts$n, p = rep(1/7, 7))
print(chi_test)

# Effect size
cramers_v <- sqrt(chi_test$statistic / (sum(monthly_counts$n) * (7-1)/7))
cat("Effect size (Cramer's V):", round(cramers_v, 3), "\n")
# Measures how strong the monthly differences are (0 = no effect, 1 = maximum effect)
# Adjusts for sample size and number of categories
```

**observed vs expected:**

**Month Observed Expected**

**jan 4991 4639.429 (+351.6)**

**feb 4573 4639.429 (-66.4)**

**mar 4907 4639.429 (+267.6)**

**apr 4407 4639.429 (-232.4)**

**may 4792 4639.429 (+152.6)**

**jun 4811 4639.429 (+171.6)**

**jul 3995 4639.429 (-644.4)**

What the Chi-Square Test Does: It calculates how much your observed
counts (actual data) differ from these expected counts (equal
distribution). The large differences (especially January and July) are
what produced your significant p-value.

Statistical Conclusion:

REJECT the Null Hypothesis Interpretation:

p \< 0.05: Reject H₀ - Significant evidence that flight frequencies vary
by month p ≥ 0.05: Fail to reject H₀ - No significant evidence of
monthly variation

Interpretation:

p-value \< 2.2e-16 (extremely small) → Strong evidence against H₀
X-squared = 155.54 with 6 degrees of freedom → Large test statistic

Result: Flight frequencies are NOT equally distributed across months

Statistical Significance: The probability of observing this uneven
distribution by random chance is virtually zero There are statistically
significant differences in flight frequencies across months

ffect size (Cramer's V): 0.075 means you have a SMALL EFFECT SIZE

Interpreting Cramer's V:

Value Range Effect Size Interpretation 0.00 - 0.06 Negligible No
practical difference 0.07 - 0.20 Small Weak association 0.21 - 0.35
Medium Moderate association 0.36+ Large Strong association

Real-World Interpretation:

While monthly differences are real and not due to chance The magnitude
of these differences is quite small Month alone explains very little of
the variation in flight frequencies

```{r}
# Check actual monthly distribution
monthly_counts %>%
  mutate(percentage = round(n/sum(n)*100, 1)) %>%
  arrange(month_name)
```

**Statistical Conclusion: REJECT NULL HYPOTHESIS ✅**

**Statistical Interpretation**

"We reject the null hypothesis that flight frequencies are uniformly
distributed across months. The Chi-square goodness of fit test reveals
statistically significant variation in flight frequencies across
different months (χ² = 155.54, df = 6, p-value = 2.2e)."


##  Research Question: How do flight frequencies and passenger volumes
differ across carrier types (major, national, and regional carriers)?
Hypotheses:

### Null Hypothesis (H₀):

There is no significant difference in flight frequencies and passenger
volumes among major carriers, national carriers, and regional carriers.

####  Statistical Form: μ_major = μ_national = μ_regional (where μ represents
mean flight frequencies or passenger volumes)

### Alternative Hypothesis (H₁):

There are significant differences in flight frequencies and passenger
volumes among carrier types, with major carriers handling higher volumes
than national and regional carriers. Statistical Form: At least one
carrier type has significantly different mean flight frequencies or
passenger volumes, with major carriers expected to demonstrate the
highest values.

### Variables:

####  Response variables: flight_count (numerical), passengers (numerical)
####  Explanatory variable: carrier_type (categorical, 3 levels: major,
national, regional)

```{r}
# Statistical Analysis: Carrier Type vs Flight Frequencies
cat("\n=== CARRIER TYPE ANALYSIS ===\n")

# 1. Count flights by carrier type
carrier_summary <- flights %>%
  group_by(carrier_type) %>%
  summarise(
    flight_count = n(),
    total_passengers = sum(passengers),
    avg_passengers = mean(passengers),
    .groups = 'drop'
  )

print(carrier_summary)

```

```{r}
# 2. Visualize the differences
ggplot(carrier_summary, aes(x = carrier_type, y = flight_count, fill = carrier_type)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = flight_count), vjust = -0.5, size = 5) +
  labs(title = "Flight Counts by Carrier Type",
       x = "Carrier Type", y = "Number of Flights") +
  theme_classic()
```

```{r}
# 3. Perform Chi-Square Test (Simplest approach)
cat("\n=== CHI-SQUARE TEST ===\n")
chi_test <- chisq.test(carrier_summary$flight_count)
print(chi_test)
```

**Statistical Conclusion: REJECT THE NULL HYPOTHESIS**

Interpretation:

p-value \< 2.2e-16: Extremely significant (virtually zero chance this
pattern is random) X-squared = 11,355: Massive test statistic df = 2:
Comparing 3 carrier types

What This Means:

STRONG EVIDENCE that flight frequencies are NOT equally distributed
across carrier types

```{r}
# 4. Compare passenger volumes (ANOVA)
cat("\n=== PASSENGER VOLUME COMPARISON ===\n")
anova_test <- aov(passengers ~ carrier_type, data = flights)
summary(anova_test)
```

**Statistical Conclusion: REJECT THE NULL HYPOTHESIS**

**ANOVA Results Interpretation:**

p-value \< 2e-16: Extremely significant F-value = 615.1: Very large
F-statistic **Significance:** \*: Highest significance level What This
Means:

STRONG EVIDENCE that passenger volumes differ significantly across
carrier types

**Key Findings:**

Flight counts differ by carrier type (Chi-square test) Passenger volumes
differ by carrier type (ANOVA test)

```{r}
library(kableExtra)
# 5. Simple percentage breakdown
cat("\n=== PERCENTAGE BREAKDOWN ===\n")
carrier_summary %>%
  mutate(
    flight_percent = round(flight_count / sum(flight_count) * 100, 1),
    passenger_percent = round(total_passengers / sum(total_passengers) * 100, 1)
  ) %>%
  select(carrier_type, flight_count, flight_percent, total_passengers, avg_passengers, passenger_percent)%>%
  kable()

```

**Key Insight:**

Major carriers dominate the market, handling:

61% of all flights 58% of all passengers

```{r}
# Create summary data for analysis
flight_summary <- flights %>%
  group_by(month, carrier_type) %>%
  summarise(flight_count = n(), .groups = 'drop')

# View the summary data
print(flight_summary)
```

```{r}
# Create the main visualization
ggplot(flight_summary, aes(x = factor(month), y = flight_count, fill = carrier_type)) +
  geom_col(position = "dodge") +
  labs(title = "Flight Frequencies by Month and Carrier Type",
       subtitle = "Analysis of Seasonal Patterns Across Different Carrier Types",
       x = "Month",
       y = "Flight Count",
       fill = "Carrier Type") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(legend.position = "bottom")

# Alternative: Line plot to better show trends
ggplot(flight_summary, aes(x = month, y = flight_count, color = carrier_type, group = carrier_type)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  labs(title = "Seasonal Flight Frequency Trends by Carrier Type",
       subtitle = "Monthly variation in flight operations",
       x = "Month",
       y = "Flight Count",
       color = "Carrier Type") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(breaks = 1:7) +
  theme_minimal()
```

```{r}
# Calculate percentage distribution
monthly_totals <- flight_summary %>%
  group_by(month) %>%
  summarise(total_flights = sum(flight_count))

flight_percentage <- flight_summary %>%
  left_join(monthly_totals, by = "month") %>%
  mutate(percentage = (flight_count / total_flights) * 100)

# View percentage distribution
print(flight_percentage)
```

```{r}
#Perform Chi-Square Test (Simplest approach)
cat("\n=== CHI-SQUARE TEST ===\n")
chi_test <- chisq.test(flight_summary$flight_count)
print(chi_test)
```

#### Total Flights by month

```{r}
# Calculate total flights by month
monthly_flights <- flights %>%
  group_by(month) %>%
  summarise(
    total_flights = n(),
    .groups = 'drop'
  ) %>%
  arrange(desc(total_flights))

# Create month labels
month_labels <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul")

# Create proper month order while showing highest to lowest
monthly_flights_ordered <- monthly_flights %>%
  mutate(
    month_name = factor(month, 
                       levels = 1:7,
                       labels = month_labels[1:7]),
    rank = rank(-total_flights)
  ) %>%
  arrange(desc(total_flights))

# Create the plot
ggplot(monthly_flights_ordered, aes(x = reorder(month_name, -total_flights), 
                                   y = total_flights,
                                   fill = total_flights)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = paste0(format(total_flights, big.mark = ","), 
                              "\n(", rank, ")")), 
            vjust = -0.08, size = 1.9, fontface = "bold") +
  scale_fill_gradient(low = "lightblue", high = "darkblue", 
                     name = "Total Flights") +
  labs(title = "monthly Flight Volume (January - July)",
       subtitle = "Ranked from highest to lowest, with rank numbers in parentheses",
       x = "month",
       y = "Total Number of Flights") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 10),
        plot.subtitle = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 0, hjust = 0.5))

```

```{r}
# Correlation between distance and month
cor(monthly_flights$month, monthly_flights$total_flights, use = "complete.obs")
```

There is a strong correlation between month and total flights

#### Passenger by carrier


```{r}
# 6. ANOVA to test categorical relationships
cat("\n=== ANOVA: PASSENGERS BY CARRIER TYPE ===\n")
anova_carrier <- aov(passengers ~ carrier_type, data = flights)
summary(anova_carrier)

cat("\n=== ANOVA: PASSENGERS BY DISTANCE CATEGORY ===\n")
anova_distance <- aov(passengers ~ distance_cat, data = flights)
summary(anova_distance)

```

## LINEAR REGRESSION FOR PASSENGERS

```{r}
# linear regression

better_model <- lm(passengers ~ distance + carrier_type + month, data = flights)
summary(better_model)
```

```{r}
# linear regression with distance only
model <- lm(passengers ~ carrier_type, data = flights)
summary(model)
```

```{r}
# linear regression with distance only
model <- lm(passengers ~ distance, data = flights)
summary(model)
```

```{r}
# linear regression with distance AND carrier_type
model <- lm(passengers ~ distance + carrier_type, data = flights)
summary(model)
```

```{r}
# linear regression with distance only
model <- lm(passengers ~ distance+ carrier_type+ month, data = flights)
summary(model)
```

```{r}
# linear regression
model <- lm(passengers ~ distance + carrier_type , data = flights)
summary(model)
```

```{r}
model <- lm(passengers ~ distance * carrier_type , data = flights)
summary(model)
```

```{r}
# Create aggregated features
origin_features <- flights %>%
  group_by(origin_city_name) %>%
  summarise(
    origin_avg_passengers = mean(passengers),
    origin_flight_count = n(),
    .groups = "drop"
  )

dest_features <- flights %>%
  group_by(dest_city_name) %>%
  summarise(
    dest_avg_passengers = mean(passengers),
    dest_flight_count = n(),
    .groups = "drop"
  )

# Merge and use
flights_features <- flights %>%
  left_join(origin_features, by = "origin_city_name") %>%
  left_join(dest_features, by = "dest_city_name")

model_features <- lm(passengers ~ distance + carrier_type + 
                       origin_avg_passengers + dest_avg_passengers, 
                     data = flights_features)
summary(model_features)
```

## Linear regression for flights

```{r}
# Your aggregated data is ready
flight_counts <- flights %>%
  group_by(carrier_type, month) %>%
  summarise(
    flight_count = n(),
    total_passengers = sum(passengers),
    avg_passengers = mean(passengers),
    .groups = "drop"
  )

# 1. Simple carrier type model
model1 <- lm(flight_count ~ carrier_type, data = flight_counts)
summary(model1)

# 2. Add month (now meaningful with aggregated data)
model2 <- lm(flight_count ~ carrier_type + month, data = flight_counts)
summary(model2)

# 3. Interaction model
model3 <- lm(flight_count ~ carrier_type * month, data = flight_counts)
summary(model3)

# 4. Visualize
library(ggplot2)

# Plot 1: Flight counts by carrier
p1 <- ggplot(flight_counts, 
             aes(x = carrier_type, y = flight_count, fill = carrier_type)) +
  geom_boxplot() +
  labs(title = "Flight Counts by Carrier Type",
       x = "Carrier Type", y = "Number of Flights") +
  theme_minimal()

# Plot 2: Monthly trends by carrier
p2 <- ggplot(flight_counts, 
             aes(x = month, y = flight_count, color = carrier_type, 
                 group = carrier_type)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(title = "Monthly Flight Trends by Carrier",
       x = "Month", y = "Number of Flights") +
  theme_minimal()

#print(p1)
#print(p2)
```

```{r}
# Create aggregated data with both flight counts and passenger info
flight_data <- flights %>%
  group_by(carrier_type, month) %>%  # grouping variables
  summarise(
    flight_count = n(),               # Number of flights
    total_passengers = sum(passengers),  # Total passengers
    avg_passengers = mean(passengers),   # Average per flight
    .groups = "drop"
  )
print(summary(flight_data))
# Does passenger volume predict flight frequency?
model1 <- lm(flight_count ~ total_passengers, data = flight_data)
summary(model1)

# Or average passengers
model2 <- lm(flight_count ~ avg_passengers, data = flight_data)
summary(model2)

# Flight count predicted by carrier type AND passengers
model3 <- lm(flight_count ~ carrier_type + avg_passengers, data = flight_data)
summary(model3)

# Most comprehensive model
best_model <- lm(flight_count ~ carrier_type + avg_passengers + month, 
                 data = flight_data)
summary(best_model)

```

```{r}
# Model 1: Flight count by carrier type
model1 <- lm(flight_count ~ carrier_type, data = flight_counts)
summary(model1)

# Model 2: Flight count by month (seasonality)
model2 <- lm(flight_count ~ month, data = flight_counts)  
summary(model2)

# Model 3: Both predictors
model3 <- lm(flight_count ~ carrier_type + month, data = flight_counts)
summary(model3)

```
