---
title: "Final Project 606 Visualizations"
author: "Mehreen Ali Gillani"
output:
  html_document: default
  pdf_document: default
date: "2025-11-26"
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# import Libraries

library(GGally)
library(scales)
library(tidyverse)
library(corrplot)
library(gridExtra)
library(patchwork)
library(ggcorrplot)
library(viridis)
```

# Flight Data Analysis: U.S. Domestic Air Travel Patterns (Jan-July 2025)

## Project Overview

This analysis examines U.S. domestic air travel patterns during the
first seven months of 2025, utilizing data from the T-100 Domestic
Market (U.S. Carriers) dataset. The comprehensive visualization explores
key metrics including passenger volume, flight operations, freight, mail
transportation, and distance patterns across the national air network.

## Analysis Scope

The investigation employs extensive univariate and multivariate data
visualization techniques to uncover insights about:

-   Passenger distribution and travel demand fluctuations

-   Flight frequency patterns across major routes

-   Origin-destination dynamics and popular travel corridors

-   Carrier performance comparisons across three categories: Major,
    National, and Regional carriers

## Key Focus Areas

Through correlation analysis and comparative visualization, this report
specifically examines relationships between:

-   Passenger volume and flight frequency

-   Distance traveled and carrier type preferences

-   Geographic patterns in origin and destination cities

-   Market segmentation across different carrier classifications

## Methodology

The analysis employs R programming with visualization packages including
ggplot2 to create interactive and informative representations of complex
air travel data, providing stakeholders with actionable insights into
2025 domestic aviation trends.

**Read Flights**

```{r read_file}

url <- "https://raw.githubusercontent.com/mehreengillani/Final_project_Data606/refs/heads/main/flights_data_clean.csv"
flights <- read_csv(url)

```

```{r}
# Verify the new dataframe
dim(flights)
colnames(flights)
view(flights)
```

## 1. UNI-variate analysis:

#### 1.1 NUMERICAL VARIABLES

```{r}
# Set theme for better visuals
theme_set(theme_minimal())

# Passengers distribution
p1 <- ggplot(flights, aes(x = passengers)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  labs(title = "Distribution of Passengers",
       x = "Number of Passengers", y = "Count") +
  scale_x_continuous(labels = scales::comma)

# Distance distribution
p2 <- ggplot(flights, aes(x = distance)) +
  geom_histogram(bins = 30, fill = "darkorange", alpha = 0.7) +
  labs(title = "Distribution of Distance",
       x = "Distance", y = "Count") +
  scale_x_continuous(labels = scales::comma)

# 1. Create binary indicator
flights <- flights %>%
  mutate(has_cargo = ifelse(mail > 0 | freight > 0, "Yes", "No"))

# Add title to the table output
cat("=== CARGO VS PASSENGER-ONLY FLIGHTS ===\n")
table(flights$has_cargo)

# Or with proportions
prop.table(table(flights$has_cargo))
# 2. Plot cargo vs no-cargo flights
p3 <- ggplot(flights, aes(x = has_cargo, fill = has_cargo)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count..), 
            vjust = -0.5, size = 4, fontface = "bold") +
  labs(title = "Flights with Cargo vs Passenger-Only",
       x = "Carries Cargo?", y = "Number of Flights")

# 3. Distribution of non-zero cargo
cargo_flights <- flights %>% filter(mail > 0 | freight > 0)

# mail distribution
p4 <-ggplot(cargo_flights, aes(x = mail)) +
  geom_histogram(bins = 30, fill = "brown") +
  labs(title = "Mail Distribution \n (Non-zero flights only)")

# Freight distribution
p5 <- ggplot(cargo_flights, aes(x = freight)) +
  geom_histogram(bins = 30, fill = "darkred") +
  labs(title = "Freight Distribution\n(Non-zero flights only)")
# Numerical plots grid
grid.arrange(p1, p2, p5, p4, ncol = 2)
print(p3)
```

#### 1.2 CATEGORICAL VARIABLES

```{r}

# Carrier type distribution
p5 <- ggplot(flights, aes(x = carrier_type)) +
  geom_bar(fill = "skyblue", alpha = 0.7) +
  labs(title = "Distribution by Carrier Type",
       x = "Carrier Type", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Month distribution
p6 <- ggplot(flights, aes(x = month_name)) +
  geom_bar(fill = "lightgreen", alpha = 0.7) +
  geom_text(stat = 'count', aes(label = ..count..), vjust = -0.2) +
  labs(title = "Distribution by Month",
       x = "Month", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Origin city distribution (top 15)
top_origins <- flights %>%
  count(origin_city_name) %>%
  arrange(desc(n)) %>%
  head(15)

p7 <- ggplot(top_origins, aes(x = reorder(origin_city_name, n), y = n)) +
  geom_bar(stat = "identity", fill = "coral", alpha = 0.7) +
  labs(title = "Top 15 Origin Cities",
       x = "Origin Cities", y = "Count") +
  coord_flip()

# Destination city distribution (top 15)
top_dests <- flights %>%
  count(dest_city_name) %>%
  arrange(desc(n)) %>%
  head(15)

p8 <- ggplot(top_dests, aes(x = reorder(dest_city_name, n), y = n)) +
  geom_bar(stat = "identity", fill = "goldenrod", alpha = 0.7) +
  labs(title = "Top 15 Destination Cities",
       x = "Destination Cities", y = "Count") +
  coord_flip()

# Distance category distribution
p99 <- ggplot(flights, aes(x = distance_cat)) +
  geom_bar(fill = "lightpink", alpha = 0.7) +
  labs(title = "Distribution by Distance Category",
       x = "Distance Category", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Create new distance categories (5 categories)
flights_distance <- flights %>%
  mutate(distance_cat_5 = cut(distance,
                             breaks = c(0, 300, 600, 1200, 2400, Inf),
                             labels = c("Short(0-300mi)", 
                                        "Medium-Short(301-600mi)",
                                        "Medium(601-1200mi)", 
                                        "Long(1201-2400mi)",
                                        "Very Long(2400+mi)"),
                             include.lowest = TRUE))

# Plot with new categories
p9 <- ggplot(flights_distance, aes(x = distance_cat_5)) +
  geom_bar(fill = "#2E86AB") +
  geom_text(stat = 'count', aes(label = ..count..), vjust = -0.2) +
  labs(title = "Flights by Distance", x = "Distance", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


# Class distribution
p10 <- ggplot(flights, aes(x = class)) +
  geom_bar(fill = "lightsteelblue", alpha = 0.7) +
  labs(title = "Distribution by Class",
       x = "Class", y = "Count")


# Get top 15 routes
top_routes <- flights %>%
  count(route) %>%
  arrange(desc(n)) %>%
  head(15)

# Plot top 15 routes
p11 <- ggplot(top_routes, aes(x = reorder(route, n), y = n)) +
  geom_bar(stat = "identity", fill = "lightblue", alpha = 0.8) +
  geom_text(aes(label = scales::comma(n)), 
            hjust = -0.2, size = 3.5, color = "darkblue") +
  labs(title = "Top 15 Busiest Routes",
       subtitle = "Most frequently traveled flight routes",
       x = NULL, 
       y = "Number of Flights") +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.1))) +
  coord_flip() 


# 3. ARRANGE PLOTS IN GRID

# Arrange plots with proper spacing and sizing
grid.arrange(p5, p6, ncol = 2)
             #padding = unit(3, "cm")
grid.arrange(p9,p10, ncol = 2)
# State distribution plots
grid.arrange(p7, p8, ncol = 2)

#Route
grid.arrange(p11, ncol = 2)
```

```{r}
# summary for key metrics only
summary_table <- flights %>% 
  select(carrier_type, month_name, class) %>%
  summary() 
print(summary_table)
summary(flights_distance$distance_cat_5)
```

```{r}
# Summary statistics for numerical variables
summary_stats <- flights %>%
  select(passengers, distance, freight, mail) %>% #,class, carrier_type, month_name, distance_group) %>%
  summary()
print(summary_stats)

```

## 2. Multi-variate Analysis

```{r}
# Scatter plot with alpha for overplotting
p1 <- ggplot(flights, aes(x = distance, y = passengers)) +
  geom_point(alpha = 0.3, size = 1, color = "darkblue") +
  labs(title = "Passengers vs Distance",
       subtitle = "Scatter plot showing relationship between flight distance and passenger count",
       x = "Distance (miles)", 
       y = "Number of Passengers") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(labels = scales::comma) +
  theme(plot.title = element_text(face = "bold", size = 14))

print(p1)

```

```{r}
# Hexbin plot for dense data
p2 <- ggplot(flights, aes(x = distance, y = passengers)) +
  geom_hex(bins = 50) +
  scale_fill_viridis_c(name = "Number of\nFlights") +
  labs(title = "Passengers vs Distance",
       subtitle = "Hexbin plot showing density of flights",
       x = "Distance (miles)", 
       y = "Number of Passengers") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(labels = scales::comma) +
  theme(plot.title = element_text(face = "bold", size = 14))

print(p2)
```

```{r}
# Set theme
theme_set(theme_minimal())

## PLOT 1: Passengers vs Distance colored by Carrier Type
p1 <- ggplot(flights, aes(x = distance, y = passengers, color = carrier_type)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_viridis_d(name = "Carrier Type") +  # Changed to _d for discrete
  labs(title = "Passengers vs Distance by Carrier Type",
       subtitle = "Relationship between flight distance and passenger count across different carrier types",
       x = "Distance (miles)", 
       y = "Number of Passengers") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(labels = scales::comma) +
  theme(plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 10, color = "gray50"),
        legend.position = "bottom") +
  guides(color = guide_legend(nrow = 2, byrow = TRUE))

print(p1)
```

```{r}
## PLOT 2: Freight vs Distance by Carrier Type
p2 <- ggplot(flights, aes(x = distance, y = freight, color = carrier_type)) +
  geom_point(alpha = 0.7, size = 1.5) +
  labs(title = "Freight vs Distance by Carrier Type",
       subtitle = "Relationship between flight distance and freight volume across different carrier types",
       x = "Distance (miles)", 
       y = "Freight Volume") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(labels = scales::comma) +
  theme(plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 10, color = "gray50"),
        legend.position = "bottom") +
  guides(color = guide_legend(nrow = 2, byrow = TRUE))

print(p2)
```

```{r}
# Distance Distribution by Month
p1 <- ggplot(flights, aes(x = factor(month), y = distance, fill = factor(month))) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Distance Distribution by Month",
       subtitle = "Boxplot showing distance variations across months",
       x = "Month", y = "Distance (miles)") +
  scale_fill_viridis_d() +
  guides(fill = "none") +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))

print(p1)
```

```{r}
# Line plot with confidence intervals showing monthly trends
p3 <- flights %>%
  group_by(month) %>%
  summarise(
    avg_distance = mean(distance, na.rm = TRUE),
    se_distance = sd(distance, na.rm = TRUE) / sqrt(n())
  ) %>%
  ggplot(aes(x = month, y = avg_distance)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_ribbon(aes(ymin = avg_distance - 1.96*se_distance, 
                  ymax = avg_distance + 1.96*se_distance), 
              alpha = 0.2, fill = "steelblue") +
  labs(title = "Average Distance by Month",
       subtitle = "Line plot with confidence intervals showing monthly trends",
       x = "Month", y = "Average Distance (miles)") +
  scale_x_continuous(breaks = 1:12)+ ylim(500, NA)

print(p3)
```

```{r}
# First aggregate the data
monthly_agg <- flights %>%
  count(month, carrier_type) %>%
  mutate(month_name = factor(month.abb[month], levels = month.abb))

p2 <- ggplot(monthly_agg, 
             aes(x = month_name, y = n, 
                 color = carrier_type, group = carrier_type)) +
  geom_line(size = 1, alpha = 0.8) +
  geom_point(size = 2) +
  scale_color_brewer(palette = "Set1", name = "Carrier Type") +
  labs(title = "Monthly Flight Trends by Carrier Type",
       subtitle = "Line chart showing seasonal patterns",
       x = "Month", 
       y = "Number of Flights") +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p2)
summary(monthly_agg)

```

```{r}
# Ensure correct ordering
p3 <- flights %>%
  mutate(passenger_group = case_when(
    passengers <= 150 ~ "1. Small (≤150)",
    passengers <= 300 ~ "2. Medium (151-300)",
    passengers <= 500 ~ "3. Large (301-500)",
    TRUE ~ "4. Very Large (>500)"
  )) %>%
  group_by(month, passenger_group) %>%
  summarise(
    avg_distance = mean(distance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = month, y = avg_distance, color = passenger_group, group = passenger_group)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Average Flight Distance by Passenger Group",
       subtitle = "Grouped by passenger capacity",
       x = "Month", y = "Average Distance (miles)") +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_color_brewer(palette = "Set1", name = "Passenger Group",
                    labels = c("Small (≤150)", "Medium (151-300)", 
                              "Large (301-500)", "Very Large (>500)")) +
  theme_minimal()

print(p3)
```

```{r}
p1_labeled <- flights %>%
  mutate(passenger_group = factor(case_when(
    passengers <= 150 ~ "Small (≤150)",
    passengers <= 300 ~ "Medium (151-300)",
    passengers <= 500 ~ "Large (301-500)",
    TRUE ~ "Very Large (>500)"
  ), levels = c("Small (≤150)", "Medium (151-300)", "Large (301-500)", "Very Large (>500)"))) %>%
  count(carrier_type, passenger_group, month) %>%
  mutate(month_name = factor(month.abb[month], levels = month.abb)) %>%
  ggplot(aes(x = month_name, y = carrier_type, fill = n)) +
  geom_tile(color = "white", size = 0.5) +
  geom_text(aes(label = ifelse(n > 50, scales::comma(n), "")), 
            size = 2.8, color = "white", fontface = "bold") +
  facet_wrap(~ passenger_group, ncol = 2) +
  scale_fill_gradient(
    name = "Flights",
    low = "skyblue", 
    high = "darkblue",
    trans = "sqrt",
    labels = scales::comma
  ) +
  labs(title = "Monthly Flight Distribution by Carrier and Passenger Group",
       x = "Month", y = "Carrier Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p1_labeled)
```

```{r}
p2 <- flights %>%
  group_by(carrier_type, month) %>%
  summarise(total_passengers = sum(passengers, na.rm = TRUE), .groups = "drop") %>%
  mutate(month_name = factor(month.abb[month], levels = month.abb)) %>%
  ggplot(aes(x = month_name, y = total_passengers, fill = carrier_type)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  scale_fill_brewer(palette = "Set2", name = "Carrier Type") +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Monthly Passenger Volume by Carrier Type",
       subtitle = "Total passengers transported each month",
       x = "Month", y = "Total Passengers") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12, color = "gray50")
  )

print(p2)
```

```{r}
p4 <- ggplot(flights, aes(x = carrier_type, y = passengers, fill = carrier_type)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Passenger Distribution by Carrier Type",
       x = "Carrier Type", y = "Passengers") +
  scale_y_continuous(labels = scales::comma) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Set2")
print(p4)
```

```{r}
# Prepare the data
top_routes_10 <- flights %>%
  group_by(origin_city_name, dest_city_name) %>%
  summarise(
    total_passengers = sum(passengers, na.rm = TRUE),
    total_flights = n(),
    #avg_passengers = mean(passengers, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(total_passengers)) %>%
  head(10) %>%
  mutate(route_label = paste(origin_city_name, "→", dest_city_name))

p11_clean <- ggplot(top_routes_10, 
                    aes(x = reorder(route_label, total_passengers), 
                        y = total_passengers)) +
  geom_col(fill = "lightblue", width = 0.6, alpha = 0.8) +
  geom_text(aes(label = scales::comma(total_passengers)), 
            hjust = -0.2, size = 4.5, fontface = "bold", color = "#2E86AB") +
  geom_text(aes(label = paste("(", total_flights, "flights)")), 
            hjust = 1.1, size = 3.5, color = "darkblue") +
  labs(title = "Top 10 Busiest Air Routes",
       subtitle = "Passenger volume and flight frequency",
       x = NULL, y = NULL) +
  scale_y_continuous(labels = NULL, expand = expansion(mult = c(0, 0.25))) +
  coord_flip() +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 20, hjust = 0.5),
    plot.subtitle = element_text(size = 14, color = "darkblue", hjust = 0.5),
    axis.text.y = element_text(size = 13, face = "bold", margin = margin(r = 15)),
    panel.grid = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks = element_blank()
  )

print(p11_clean)
```

```{r}
# Get top 10 carriers by passenger volume
top_carriers <- flights %>%
  group_by(unique_carrier_name) %>%
  summarise(
    total_passengers = sum(passengers, na.rm = TRUE),
    total_flights = n(),
    avg_passengers = mean(passengers, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(total_passengers)) %>%
  head(10)

# Create the plot
p_carriers <- ggplot(top_carriers, 
                     aes(x = reorder(unique_carrier_name, total_passengers), 
                         y = total_passengers)) +
  geom_col(fill = "#4C72B0", width = 0.7, alpha = 0.8) +
  geom_text(aes(label = scales::comma(total_passengers)), 
            hjust = -0.1, size = 4, fontface = "bold", color = "#4C72B0") +
  geom_text(aes(label = paste("(", total_flights, "flights)")), 
            hjust = 1.1, size = 3.5, color = "gray40") +
  labs(title = "Top 10 Airlines by Passenger Volume",
       subtitle = "Total passengers transported with flight count",
       x = NULL, 
       y = NULL) +
  scale_y_continuous(labels = NULL, expand = expansion(mult = c(0, 0.2))) +
  coord_flip() +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
    plot.subtitle = element_text(size = 14, color = "gray50", hjust = 0.5),
    axis.text.y = element_text(size = 12, face = "bold", margin = margin(r = 15)),
    panel.grid = element_blank(),
    axis.text.x = element_blank()
  )

print(p_carriers)
```

```{r}
# Select only numeric variables for correlation analysis
numeric_vars <- flights %>%
  select(passengers, freight, mail, distance, month, distance_group)

# Correlation matrix
cor_matrix <- cor(numeric_vars, use = "complete.obs")
print(cor_matrix)

# Correlation plot
corrplot(cor_matrix, method = "color", type = "upper", 
         tl.cex = 0.8, tl.col = "black", 
         title = "Correlation Matrix of Flight Variables",
         mar = c(0,0,1,0))

# Scatterplot matrix for key variables
pairs(numeric_vars[,1:4], pch = 19, cex = 0.3, 
      main = "Scatterplot Matrix: Passengers, Freight, Mail, Distance")

# Focus on passenger relationships
library(ggplot2)
ggplot(flights, aes(x = distance, y = passengers)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Passengers vs Distance",
       subtitle = paste("Correlation:", round(cor(flights$distance, flights$passengers, use = "complete.obs"), 3)))

```

## 3. Correlation

```{r}
# 1. Convert categorical variables to numeric for 

flights_cor <- flights %>%
  mutate(
    carrier_type_num = as.numeric(as.factor(carrier_type)),
    distance_cat_num = as.numeric(as.factor(distance_cat)),
    month_num = as.numeric(as.factor(month_name)),
    origin_num = as.numeric(as.factor(origin_city_name)),
    dest_num = as.numeric(as.factor(dest_city_name)),
    class_num = as.numeric(as.factor(class))
  ) %>%
  select(passengers, freight, mail, distance,
         carrier_type_num, distance_cat_num, month_num, origin_num, dest_num, class_num)

# 2. Correlation matrix with categoricals
cor_matrix_full <- cor(flights_cor)
print(cor_matrix_full)

# 3. Visualize full correlation matrix
corrplot(cor_matrix_full, method = "color", type = "upper",
         tl.cex = 0.7, tl.col = "black",
         title = "Correlation Matrix Including Categorical Variables",
         mar = c(0,0,2,0))

```

```{r}
# Which carrier types and distance categories have the most passengers?
flights %>%
  group_by(carrier_type) %>%
  summarise(mean_passengers = mean(passengers)) %>%
  arrange(desc(mean_passengers))

flights %>%
  group_by(distance_cat) %>%
  summarise(mean_passengers = mean(passengers)) %>%
  arrange(desc(mean_passengers))
# Plot 1: Passengers by Carrier Type
p1 <- ggplot(flights, aes(x = carrier_type, y = passengers, fill = carrier_type)) +
  geom_boxplot(alpha = 0.8) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "red") +
  labs(title = "Passenger Distribution by Carrier Type",
       subtitle = "ANOVA: F = 615.1, p < 0.001 ***",
       x = "Carrier Type", y = "Number of Passengers") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  theme(legend.position = "none")

# Plot 2: Passengers by Distance Category
p2 <- ggplot(flights, aes(x = distance_cat, y = passengers, fill = distance_cat)) +
  geom_boxplot(alpha = 0.8) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "red") +
  labs(title = "Passenger Distribution by Distance Category",
       subtitle = "ANOVA: F = 93.51, p < 0.001 ***",
       x = "Distance Category", y = "Number of Passengers") +
  scale_fill_brewer(palette = "Set3") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

# Display plots
library(patchwork)
(p1 | p2) 
```

```{r}
monthly_trends <- flights %>%
  group_by(month) %>%
  summarise(
    mean_passengers = mean(passengers),
    ci_lower = mean_passengers - 1.96 * (sd(passengers) / sqrt(n())),
    ci_upper = mean_passengers + 1.96 * (sd(passengers) / sqrt(n())),
    .groups = "drop"
  ) %>%
  mutate(month_name = factor(month.abb[month], levels = month.abb))

ggplot(monthly_trends, aes(x = month_name, y = mean_passengers, group = 1)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), 
              fill = "lightblue", alpha = 0.3) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "darkblue", size = 2) +
  ylim(20, NA) +
  labs(title = "Monthly Passenger Trends with 95% CI",
       subtitle = "Shaded area shows uncertainty in monthly estimates",
       x = "Month", y = "Average Passengers")
```

```{r}
# Create aggregated data with both flight counts and passenger info
flight_data <- flights %>%
  group_by(carrier_type, month) %>%  # grouping variables
  summarise(
    flight_count = n(),               # Number of flights
    total_passengers = sum(passengers),  # Total passengers
    avg_passengers = mean(passengers),   # Average per flight
    .groups = "drop"
  )
p3 <- ggplot(flight_data, 
                   aes(x = total_passengers, y = flight_count, 
                       color = carrier_type)) +
  # Size only on points, not global
  geom_point(aes(size = avg_passengers), alpha = 0.7) +
  # No size in geom_smooth
  geom_smooth(method = "lm", se = FALSE, aes(group = carrier_type)) +
  labs(title = "Flights vs Total Passengers",
       subtitle = "Bubble size = Average passengers per flight",
       x = "Total Passengers", y = "Number of Flights") +
  scale_size_continuous(name = "Avg Passengers") +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()

print(p3)
```

## Key Findings:

-   Passenger loads are right‑skewed—most flights operate below
    capacity.

-   Cargo operations are highly specialized; many flights carry no
    freight/mail. Medium‑haul routes dominate (\>75% of flights),
    reflecting hub‑and‑spoke networks.

-   Route demand is directional: Washington ↔ Pittsburgh has asymmetric
    passenger/flight counts.

-   Carrier specialization by distance: regional (\<300 mi), national
    (mid‑haul), major (long‑haul).

-   Seasonal distance shift: short flights peak in February; longer
    routes increase in July.

-   Major carriers lead in flights and passengers across all months,
    peaking in January.

-   Regional/national carriers excel in small‑aircraft ops, not jumbo
    jets.

-   Large‑capacity flights (500+ passengers) maintain stable distances
    all year.

-   Highly fragmented market: 10,601 unique routes served by 77
    carriers.
